The role of standing variation in parallel adaptation

### 2014-09-12 standing-variation-paper.tex
Graham's changes

The issue of convergent adaptation within species touches on a number
of important questions in evolutionary biology. These include the extent
to which adaptation is shaped by pleiotropic constraints \citep{Haldane:book,Orr:05},
whether adaptation is mutation-limited \citep{karasov2010}, and whether
species should be regarded to cohesive units. Convergence within
populations also effects our
ability to detect adaptation from population genomic data as no one
allele sweeps fixation over the entire area affected by a selection
pressure \citep{softsweepsIII}.

@@

Geographically limited dispersal also increases the chance of
convergent evolution. For example when selection pressures are
geographically patchy this can increasing the probability of
convergent evolution as alleles are unable to spread through intervening populations \citep{RalphCoop:14}.
In \citep{RalphCoop} we formulated a simple model of parallel adaptation in a geographic setting
where the selection pressure was constant across the entire environment, 
and that a single mutational change was sufficient to adapt populations to the novel environment. 
We assumed that there was no standing variation for the adaptive
allele, such that parallel mutation must be due to multiple mutations
after the environmental shift. In that setting we found that there was
a characteristic geographic scale, over which we would expect multiple
instances of our adaptive allele to have arisen in parallel. This
characteristic length could be expressed in terms of a simple compound
parameter determined by our parameters of interest.

Here we extend our spatial model from \citep{RalphCoop} to include standing
variation present at mutation-selection balance before selection
pressures switch.

@@

Somewhat along similar lines, Futuyma has argued that much of the adaptive differentiation,
e.g. adaptation to local conditions, within species maybe ephemeral and
subject to loss due to local extinction and the mixing following the collapse of population
structure \citep{Futuyma:10,FUTUYMA:87}. In part this represented a
response to punctuated equilibrium \citep{eldredgegould72}, arguing
that the observation of stasis and rapid anagensis associated with
speciation were consistent with micro-evolution. 
Futuyma argued that despite rapid adaptation over short time-scales, we
may observe morphological stasis in the
fossil record as
much of this adaptation is lost \citep[see
also][]{Eldredge:05}. Furthermore, he suggests that speciation
may act as rachet to prevent the loss of differentiation, acting to
maintain adaptive changes among populations, and prevent their loss by
interbreeding. At face value the rate of species formation seems too low to
contribute to this process. However, as \citep{Rosenblum:12}, and many
others, have argued that in fact the rate of speciation may well be very high,
but that the majority of incipient species do not persist long due to
reabsorption or extinction. Repeated bouts of extinction and
speciation will send waves of alleles
to fixation along particular lineages.

%Venditti:10, 
This link between speciation and substitution does not imply that
substitutions should be thought of as being clustered at splits in
inferred phylogenies \citep[see ][for a recent exhange on
this]{Pennell:14,Venditti:14,Pennell:14B}. 
(Note that neutral substitutions will be to be unaffected by this process, as they
will accumulate in a clock like manner along lineages, as dictated by
the mutation rate, regardless of the geographic details of their
polymorphic stage.)
First, it it likely that splits in
phylogenies are likely only a tiny proportion of all incipient
speciation events because extinction rates may be high
\citep{Rosenblum:12}, such that every lineage has likely past through
many ``speciation events'' in addition to the observed ones.
Second, under this view, spatial polymorphisms may accumulate gradually, in
geographically restricted populations, between large-scale 
biogeographic events leading to their fixation (or indeed loss). 
Such that the origination process of alleles and their fixation process
are reasonably uncoupled \citep[in the sense of ][]{Gillespie:94} by the geographic structure.

>>>>

The issue of convergent adaptation within species touches on a number
of important questions in evolutionary biology. These include the extent
to which adaptation is shaped by pleiotropic constraints \citep{Haldane:book,Orr:05},
whether adaptation is mutation-limited \citep{Bradshaw:91,karasov2010}, and whether
species should be regarded to cohesive units. Convergence evolution within
populations also effects our ability to detect adaptation from population genomic data as no one
allele sweeps fixation over the entire area affected by a selection
pressure \citep{softsweepsIII}.

@@

Geographically limited dispersal also increases the chance of
convergent evolution. For example when selection pressures are
geographically patchy this can increasing the probability of
convergent evolution as alleles are unable to spread through
intervening populations \citep{RalphCoop:14}. In \citep{RalphCoop} we formulated a simple model of parallel adaptation in a geographic setting
where the selection pressure was constant across the entire environment, 
and that a single mutational change was sufficient to adapt populations to the novel environment. 
We assumed that there was no standing variation for the adaptive
allele, such that parallel mutation must be due to multiple mutations
after the environmental shift. In that setting we found that there was
a characteristic geographic scale, over which we would expect multiple
instances of our adaptive allele to have arisen in parallel. This
characteristic length could be expressed in terms of a simple compound
parameter determined by our parameters of interest. Here we extend
this spatial model include standing
variation present at mutation-selection balance before selection
pressures switch.

@@

Along somewhat similar lines, Futuyma has argued that much of the adaptive differentiation,
e.g. adaptation to local conditions, within species maybe ephemeral and
subject to loss due to local extinction and the mixing following the collapse of population
structure \citep{Futuyma:10,FUTUYMA:87}. In part this represented a
response to the pattern of punctuated equilibrium \citep{eldredgegould72}, arguing
that the observation of stasis and rapid anagensis associated with
speciation were consistent with micro-evolution \citep[see also ][]{Futuyma:1989editedbook}. 
Futuyma argued that despite rapid adaptation over short time-scales, we
may observe morphological stasis in the
fossil record as
much of this adaptation is lost \citep[see
also][]{lieberman:96,Eldredge:05}. Furthermore, he suggests that speciation
may act as rachet to prevent the loss of differentiation, acting to
maintain adaptive changes among populations, and prevent their loss by
interbreeding. At face value the rate of species formation seems too low to
contribute to this process. However, as \citep{Rosenblum:12}, and many
others, have argued that in fact the rate of speciation may well be very high,
but that the majority of incipient species do not persist long due to
reabsorption or extinction. Repeated bouts of extinction and
speciation will send waves of alleles
to fixation along particular lineages.

While Futuyma and others have advanced these ideas mainly in the
setting of adaptation to local conditions \citep[particularly local
communities ][]{} . Much of the recent adaptation that we see as population
geneticists is likely adaptation to local conditions, but as we have
shown even wide-spread selection pressures will likely result in
alleles only spreading locally. The adaptive substitution
we see along lineages are the alleles that were lucky enough to
have spread in populations that escaped local extinction, and to have
avoided being assorted into species that eventually went extinct. 
Range sizes change rapidly enough that it seems likely that range size
changes should change substantially over the soujourn time of most
alleles that will eventually become fixed in the species, as such it
seems as though .

%Venditti:10, 
This link between speciation and substitution does not imply that
substitutions should necessarily be thought of as being clustered at splits in
inferred phylogenies \citep[see ][for a recent exhange on
this]{Pennell:14,Venditti:14,Pennell:14B}. 
We first note that neutral substitutions will be to be unaffected by this process, because they
will accumulate in a clock like manner along lineages, as dictated by
the mutation rate, regardless of the geographic details of their
polymorphic stage. Turning to the accumulation of adaptive
substitutions, 
it likely that splits in
phylogenies are likely only a tiny proportion of all incipient
speciation events because extinction rates may be high
\citep{Rosenblum:12}, such that every lineage has likely past through
many ``speciation events'' in addition to the observed ones.
Futhermore, under this view, spatial polymorphisms may accumulate gradually, in
geographically restricted populations, between large-scale 
biogeographic events leading to their fixation (or indeed loss). 
Such that the origination process of alleles and their fixation process
are reasonably uncoupled \citep[in the sense of ][]{Gillespie:94} by the geographic structure.
